Equivalence Testing and the Second Generation P-Value

### 2018-08-24 manuscript.Rmd
Last typo's and rounding mistakes

There are 4 situations where *p*-values from TOST and SGPV are unrelated. The first two situations were discussed before and can be seen in Figure \ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1 *p*-values from the equivalence test fall between 0.975 and 1 or between 0 and 0.025. Because *p*-values approach 0 or 1, but are never exactly 0 or 1, and the SGPV is exactly 0 or 1, the two statistics are completely unrelated within these ranges. The easiest way to see this is by plotting the SGPV against the *p*-value from the TOST procedure. The situations where the SPGV and *p*-values from the TOST procedure are unrelated are indicated by the parts of the curve where there are vertical straight lines at second generation *p*-values of 0 and 1.
```{r, include=FALSE}
step = 0.01

@@

To conclude this section, there are four situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely continuous descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.629 to 0.629, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.992.

The effect of assymetric confidence intervals is most easily noticable at smaller sample sizes, therefore in Figure \ref{fig:TOSTSGPV11} below we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.

```{r, include=FALSE}
n <- 30
r_pop <- .0
low_eqbound_r <- -.45
high_eqbound_r <- .45
alpha <- .05
step <- 0.001

@@

This example illustrates the difference between a proportion and a probability. There is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answer, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at these values. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis'. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

Even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but in any specific case it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) is to examine whether the SGPV is 0, 1, or inconclusive. This interpretation of a SGPV as allowing researchers to reject the null, reject the presence of a meaningful effect, or remaining inconclusive is very similar to the Neyman-Pearson interpretation of combining a null-hypothesis test and an equivalence test (@lakens_equivalence_2018). The difference is that where a SGPV of 1 can be interpreted as *p* < .025, equivalence tests provide exact *p*-values, and they continue to differentiate between for example p = 0.048 and p = 0.002. Whether this is desireable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not $p < \alpha$, and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact *p*-values [@american_psychological_association_publication_2010], and exact *p*-values might provide information of interest to readers about how precisely how surprising the data is under the null model. Equivalence tests combined with null-hypothesis significance tests also allow researchers to conclude an effect is significant *and* equivalent (i.e., statistically different from zero, but also too small to be considered meaningful). Thus, the SGPV is used to classify results into one of three possible cateogies (with the data falling inside or outside the equivalence range, or being inconclusive), while equivalence tests combined with null-hypothesis tests classify results into four possible categories

@@

@blume_second-generation_2018 claim that "Adjustments for multiple comparisons are obviated" (p. 15) and that "second-generation *p*-values provide a proper scientific adjustment for multiple comparisons". However, this is not correct. With a sufficient number of looks error rates can rise above the nominal alpha level (both when compared to equivalence and minimal effect tests, but even when compared to a null-hypothesis test). When directly comparing equivalence tests and the SGPV, multiple comparisons inflate the probability that one erronously concludes there is *no* effect, where there *is* a true effect size that equals the equivalence bound, just as quickly for the SGPV as for equivalence tests, because they are directly related. To conclude, the idea that the SGPV improves error rates does not hold up under closer scrutiny, and the recommendation to ignore adjustments for multiple comparisons has the potential to increase false positives in the literature. Equivalence tests provide and easier and more formal way to control both Type I error rates (by setting the alpha level) and the Type II error rate (by performing an a-priori power analysis).

>>>>

There are three situations where *p*-values from TOST and SGPV are unrelated. The first two situations were discussed before and can be seen in Figure \ref{fig:TOSTSGPV1}. When the SGPV is either 0 or 1 *p*-values from the equivalence test fall between 0.975 and 1 or between 0 and 0.025. Because *p*-values approach 0 or 1, but are never exactly 0 or 1, and the SGPV is exactly 0 or 1, the two statistics are completely unrelated within these ranges. The easiest way to see this is by plotting the SGPV against the *p*-value from the TOST procedure. The situations where the SPGV and *p*-values from the TOST procedure are unrelated are indicated by the parts of the curve where there are vertical straight lines at second generation *p*-values of 0 and 1.
```{r, include=FALSE}
step = 0.01

@@

To conclude this section, there are three situations where the *p*-value from the TOST procedure is unrelated to the SGPV. In all these situations the *p*-value for the equivalence test differentiates tests with different means, but the SGPV does not. Therefore, as a purely continuous descriptive statistic, the SGPV is more limited than the *p*-value from the TOST procedure.

@@

So far we have only looked at the relation between equivalence tests and the SGPV when confidence intervals are symmetric (e.g., for confidence intervals around mean differences). For correlations, which are bound between -1 and 1, confidence intervals are only symmetric for a correlation of exactly 0. The confidence interval becomes increasingly asymmetric as the observed correlation nears -1 or 1. For example, with ten observations, an observed correlation of 0 has a symmetric 95% confidence interval ranging from -0.630 to 0.630, while and observed correlation of 0.7 has an asymmetric 95% confidence interval ranging from 0.126 to 0.993.

The effect of asymmetric confidence intervals is most easily noticable at smaller sample sizes, therefore in Figure \ref{fig:TOSTSGPV11} below we plot the *p*-values from equivalence tests and the SGPV (again plotted as 1-SGPV for ease of comparison) for correlations. The sample size is 30 pairs of observations, and the lower and upper equivalence bounds are set to -0.45 and 0.45, with an alpha of 0.05. As the observed correlation in the sample moves from -1 to 0 the *p*-value from the equivalence test becomes smaller, as does 1-SGPV. The pattern is quite similar to that in Figure \ref{fig:TOSTSGPV2}. The *p*-value for the TOST procedure and 1-SGPV are still identical when *p*-values are 0.975 and 0.025 (indicated by the upper and lower horizontal dotted lines). There are two important differences, however. First of all, the SGPV is no longer a straight line, but a curve, due to the asymmetry in the 95% CI. Second, and most importantly, the *p*-value for the equivalence test and the SGPV do no longer overlap at *p* = 0.5.

```{r, include=FALSE}
n <- 30
r_pop <- 0
low_eqbound_r <- -.45
high_eqbound_r <- .45
alpha <- .05
step <- 0.001

@@

This example illustrates the difference between a proportion and a probability. There is always a 50% probability of observing a correlation smaller or larger than the true correlation, but the SGPV for this situation depends on how far away the observed correlation is from 0. The further away from 0, the larger the SGPV when the observed mean falls on the equivalence bound. The SGPV is the proportion of values in a 95% confidence interval that overlap with the equivalence range, but not the probability that these values will be observed. In the most extreme case (i.e., a sample size of 4, and equivalence bounds set to r = -0.99 and 0.99, with an observed correlation of 0.99) `r round(sgpv,3)*100`% of the confidence interval overlaps with the equivalence range, even though in the long run only 50% of the correlations observed in the future will fall in this range. It should be noted that in larger sample sizes the SGPV is closer to 0.5 whenever the observed correlation falls on the equivalence bound, but this extreme example nevertheless clearly illustrates the difference between question the SGPV answers, and the question a *p*-value answers. The conclusion of this section on asymmetric confidence intervals is that a SGPV of 1 or 0 can still be interpreted as a *p* < 0.025 or *p* > 0.975 in an equivalence test, since the SGPV and *p*-value for the TOST procedure are always directly related at these values. Although @blume_second-generation_2018 state that "the degree of overlap conveys how compatible the data are with the null premise" this definition of what the SGPV provides does not hold for asymmetric confidence intervals. Although a SGPV of 1 or 0 can be directly interpreted, a SGPV between 0 and 1 is not interpretable as 'compatibility with the null hypothesis'. Indeed, Blume and colleagues write in the supplemental material that "The magnitude of an inconclusive second-generation *p*-value can vary slightly when the effect size scale is transformed. However definitive findings, i.e. a *p*-value of 0 or 1 are *not* affected by the scale changes."

@@

Even though a SGPV of 1 or 0 has a clear interpretation (we can reject effects outside or inside the equivalence range), intermediate values are not as easy to interpret (especially for effects that have asymmetric confidence intervals). In one sense, they are what they are (the proportion of overlap), but in any specific cases it can be unclear what this number tells us about the data we have collected. This is not too problematic, since the main use of the SGPV (e.g., in all examples provided by Blume and colleagues) is to examine whether the SGPV is 0, 1, or inconclusive. This interpretation of a SGPV as allowing researchers to reject the null, reject the presence of a meaningful effect, or remaining inconclusive is very similar to the Neyman-Pearson interpretation of combining a null-hypothesis test and an equivalence test (@lakens_equivalence_2018). The difference is that where a SGPV of 1 can be interpreted as *p* < .025, equivalence tests provide exact *p*-values, and they continue to differentiate between for example p = 0.048 and p = 0.002. Whether this is desirable depends on the perspective that is used. From a Neyman-Pearson perspective on statistical inferences the main conclusion is based on whether or not $p < \alpha$, and thus an equivalence test and SGPV can be performed by simply checking whether the confidence interval falls within the equivalence range, just as a null-hypothesis test can be performed by checking whether the confidence interval contains zero or not. At the same time, it is recommended to report exact *p*-values [@american_psychological_association_publication_2010], and exact *p*-values might provide information of interest to readers about how precisely how surprising the data is under the null model. Equivalence tests combined with null-hypothesis significance tests also allow researchers to conclude an effect is significant *and* equivalent (i.e., statistically different from zero, but also too small to be considered meaningful). Thus, the SGPV is used to classify results into one of three possible categories (with the data falling inside or outside the equivalence range, or being inconclusive), while equivalence tests combined with null-hypothesis tests classify results into four possible categories.

@@

@blume_second-generation_2018 claim that "Adjustments for multiple comparisons are obviated" (p. 15) and that "second-generation *p*-values provide a proper scientific adjustment for multiple comparisons". However, this is not correct. With a sufficient number of looks error rates can rise above the nominal alpha level (both when compared to equivalence and minimal effect tests, but even when compared to a null-hypothesis test). When directly comparing equivalence tests and the SGPV, multiple comparisons inflate the probability that one erronously concludes there is *no* effect, where there *is* a true effect size that equals the equivalence bound, just as quickly for the SGPV as for equivalence tests, because they are directly related. To conclude, the idea that the SGPV improves error rates does not hold up under closer scrutiny, and the recommendation to ignore adjustments for multiple comparisons has the potential to increase false positives in the literature. Equivalence tests provide an easier and more formal way to control both Type I error rates (by setting the alpha level) and the Type II error rate (by performing an a-priori power analysis).
